Composition-climate paper

### 2016-09-27 Composition_Ecology.Rmd
Changed caching to `TRUE`, minor text edits to make the paper formatting fit *Ecology Letters* style guide.

```{r setup-knitr, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  comment = " ",
  error = TRUE,
  eval = TRUE,
  cache = FALSE,
  tidy = TRUE

@@

# Effect of historic land-use, pathogens and climate change on tree-climate relationships in the northern United States

@@

* Simon J Goring: Co-conceived paper, performed analysis, developed figures and co-wrote the paper
* John "Jack" W Williams: Co-conceived paper, and co-wrote the paper

@@

* Simon J Goring
  * Department of Geography, University of Wisconsin, Madison, 550 N Park St, Madison WI 53706
* John W Williams
  * Department of Geography, University of Wisconsin, Madison, 550 N Park St, Madison WI 53706
  * Center for Climatic Research, University of Wisconsin, Madison, 1225 W Dayton St, Madison WI 53706

@@

* Main Text: 6070
* Text boxes: 0

@@

**Number of Figures, Tables & Text Boxes**: 6 (4 figures, 2 tables)

@@

Contemporary forest inventory data are widely used to understand environmental controls on tree species distributions and to project forest responses to climate change, but the stability and representativeness of contemporary tree-climate distributions is poorly understood. Here we show that tree-climate relationships for 15 common tree genera in the northern US have significantly altered over the last two centuries due to historic land use, pathogen outbreaks, and climate change. Realized niches have shifted towards higher minimum temperatures, cooler maximum temperatures, and higher rainfall. A new attribution method implicates both historic climate change and land use in these shifts, with the relative importance varying among genera and climate variables. Most climate/land-use interactions are compounding, in which historic land-use exacerbates climatic shifts toward warmer or wetter distributions. The prevalence of compounding interactions imply that contemporary-based models of species distributions may underestimate species resilience to climate change.

@@

Contemporary distributions of species, and, increasingly, paleohistorical datasets are used to map distributions of species and species traits within environmental space, study underlying processes, and prepare for climate change [@lamanna2014functional;@moritz2013future;@maguire2015modeling;@ivory2016niche;@pearman2008niche]. Distributions of species within climate space, combined with future climate scenarios and species distribution models, can be used to assess species exposure and sensitivity to climate change, identify species and habitats of concern, set conservation priorities, and prepare for shifts in habitat suitability caused by climate change  [@pellatt2012using;@iverson2013tree;@iverson1998predicting;@anderson2013nichemodels;@guisan2013predicting]. Forest inventory datasets, such as the US Forest Inventory Analysis (FIA), are particularly rich observational datasets that have been used to assess early signals of range shifts via the distributions of mature trees and seedlings [@monleon2015evidence;@zhu2012failure] and to parameterize models of current and future tree distributions [@iverson2008estimating;@iverson1998predicting].

Increasingly, ecological models and their predictive capacity are being improved by integrating contemporary observational data with paleohistorical data, when available [@moritz2013future;@maguire2015modeling]. Projections that rely only on contemporary observational data use a narrow time window that may obscure plasticity in species-climate responses [@maiorano2013building], may incompletely delimit species fundamental niches [@jackson2000responses], miss responses of species to past no-analog climates [@veloz2012no], or fail to detect effects of anthropogenic land use, and other historical legacies [@svenning2015influence]. Paleoecological records from the last deglaciation show that the time lag of forest responses to abrupt climate change varies from a few years to centuries [@williamsinpress;@webb1986vegetation;@ordonez2013realized;@svenning2015influence], suggesting that tree distributions have yet to fully respond to recent climate changes. In principle, dynamic vegetation models [*e.g.*, @leiblein2016modelling] can represent the physiological, demographic, and dispersal processes that cause tree populations to lage climate change.  In practice, however, parameterization of these models is critical, and processes operating at large temporal and spatial scales remain poorly constrained [@moorcroft2006close;@matthes2016benchmarking].

Over the last several centuries, climate change and land use have both substantially affected species distributions and vegetation structure [*e.g.*, @gehrig2007tree] and may cause species distributions to incompletely fill available climates, or otherwise be at disequilibrium [@early2014climatic;@sarmento2013impacts;@ivory2016niche]. Anthropogenic land use is highly structured and is often biased within climate space, hence, species distributions and protected natural areas often represent a biased selection from the available environmental space [@scott2001nature].  This biasing can confound studies that use contemporary distributional data to understand the environmental controls on species distributions and predict species responses to future climate scenarios [@pyke2004habitat].

Interactions between climate and land-use change can be classified into three types: (1) compounding, (2) confounding and (3) counteracting [modified from @pyke2004habitat]. Compounding effects occur when climatic and land-use change cause species distributions to shift in the same direction alog an environmental gradient.  For example, warming has been compounded by land-use conversion at the lowest elevations in the northern Sierra Nevada Mountains of California, compounding the upslope distributional shifts for butterfly species [@forister2010compounded]. Confounding effects may cause little overall change on species distributions, with land-use pressure occuring randomly across the species' environmental gradient. predictive models for Garry oak (*Quercus garryana*) using only contemporary data [@pellatt2012using] may be confounded by land-use conversion that excludes Garry oak from the deep-soil sites it formerly occupied [@macdougall2004defining;@pellatt2014environmental], but does not broadly shift distributions in climate space. Counteracting effects occur when climate and land use change act in opposite directions, narrowing the available geographic and climatic space available for species. For example, bird species in the Sierra Nevadas are expected to be pulled downslope by increasing precipitation [@tingley2012push], but this downslope migration is limited by land-use conversion in the lowlands.

@@

`r fig_nums("land_use_clim", "*(a-e) Climate change in the Upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions. Modern climate shows lower $T_{max}$, but higher $T_{min}$ relative to early-19^th^ Century normals, along with increasing $P_{ann}$.  PDSI shows strong coherence with $P_{ann}$ in this region. (f) Patterns of land-use change in the upper Midwestern United States, with the "Tension Zone", differentiating sub-boreal from southern broadleafed forests superimposed. Land use data from the NLCD [@jin2013comprehensive].*")`

Tree distributions and community composition in the upper Midwestern US have been heavily influenced by the land use conversion associated with EuroAmerican Settlement in the 1800s [@goring2016preset;@schulte2007homogenization;@hanberry2012comparison;@rhemtulla2009historical], outbreaks of exotic pathogens [@anagnostakis1987chestnut;@barnes1979stem], climate changes accompanying the end of the Little Ice Age [@hotchkiss2007response;@paquette2013climatic] and current anthropogenic warming [@wicci2011climate]. Modern forests, as represented by the Forest and Inventory Analysis surveys [@woudenberg2010forest], show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System (PLS) prior to widespread land-use conversion in the late 1800s [@goring2016preset;@schulte2007homogenization]. Wildland ecosystems have been converted to agricultural and urban land use [@rhemtulla2009historical], particularly in historical prairie and savanna ecosystems (`r fig_nums("land_use_clim", display = "cite")`f).  In northern forested regions, ranges of early successional species have expanded and abundances of late-successional species are reduced due to widespread logging in the 19^th^ and 20^th^ century, and ongoing land management [@schulte2007homogenization;@goring2016preset].  Formerly dominant tree species, *Ulmus americana* and *Castanea dentata* (at the very south-eastern edge of the domain) have experienced rangewide population collapses caused by the introduction of exotic pathogens [@barnes1979stem;@anagnostakis1987chestnut].

Here we demonstrate significant shifts in the climatic distributions of tree genera in the upper Midwest over the last 100 years as a result of historic climate change, land use, and pathogen outbreaks, and we diagnose the relative influence of historic climate and land use changes on these shifts.  We use gridded historic and contemporary climate observations (PRISM) and estimates of pre-settlement and modern forest vegetation [@goring2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwestern United States.  We construct four observed and hypothetical sets of tree-climate relationships, using a 2x2 factorial of pre-settlement and modern vegetation and climates.  We develop a new method, based on Hellinger distances among tree-climate distributions, to quantify the total change and relative importance of historic vegetation and climate change. We assess whether climate and vegetation changes are compounding, counteracting or confounding based on the magnitude and direction of species-climate shifts. We use this framework to discuss the potential impacts of historical land-use and climate change on inferences about species resilience to climate change, as predicted by species-distribution models based soley on modern distributional data.

@@

Estimates of settlement-era forest composition and tree distribution based on survey data from the PLS [@white1983history;@schulte2001original;@goring2016preset], aggregated to a 64 km^2^grid across the upper Midwestern United States [@goring2016preset]. We briefly review the methods here. PLS data originally were collected on a 1 x 1 mi grid across the Upper Midwest from 1830 - 1904 [@schulte2001original;@almendinger1996minnesota;@white1983history]. At each survey point surveyors noted the closest two to four trees and recorded their common name, distances from survey points, diameters at breast height, and azimuths to each tree. Goring et al. [-@goring2016preset] aggregated this data to an 8x8 km grid, resulting in a gridded data product with an average of 61 survey points per cell, or approximately 120 trees per cell. This process included data cleaning, taxonomic standardization, and the application of spatially varying correction factors [@goring2016preset]. Using this data, we produce presence-absence distributions for each taxon, across the upper Midwest for 15 common tree genera (`r tab_nums("genera_table", display = "cite")`). A few common names, such as "Ironwood" can not be clearly resolved to a single tree genera and so we combine *Ostrya* and *Carpinus* to Ironwood, and *Thuja* and *Juniperus* to Cedar. Here 'Presence' is defined as at least one recorded tree in any 8x8 km grid cell and 'range' is considered to be the extent in climate space of all cells with recorded presence for a specific taxon.

Modern forest data are from the US Forest Inventory and Analysis (FIA) program, which began in the 1930s and provides detailed inventories of contemporary tree distributions [@iverson2013tree;@zhu2012failure], for all regions with >10% forest cover [@woudenberg2010forest].  FIA uses a nationally standardized sampling procedure, that has evolved over time.  In the Phase 2/3 system, beginning in 1998, each plot consists of four, 7.32m radius, circular sub-plots which are sampled intensively [@woudenberg2010forest].

Comparisons between historical and modern vegetation must address differences in sampling design between the PLS and FIA datasets [@goring2016preset;@kronenfeld2014validating;@kronenfeld2010influence]. Here the primary solution for maximizing comparability is via aggregation, both spatially, by aggregating records from both datasets to a common scale (8x8 km), and taxonomically by aggregating taxa to the level of genera. Nonetheless, differences persist.  For example, the FIA contains fewer plots per 8 km cell, with more trees per plot than the PLS data. This leads to higher cell-to-cell variability in the FIA data, but these are not expected to affect the regional-scale analyses here, particularly because we are investigating presence-absence patterns. Additionally, FIA observations are collected only from forested or partially forested landscapes [@woudenberg2010forest], causing the spatial extent of many tree taxa to be lower in the FIA than in the PLS datasets (`r tab_nums("genera_table", display="cite")`). Hence, scattered trees may be present in regions without permanent or temporary FIA plots, complicating the ability to assign 'absence' within the FIA. However the practice of using FIA data as a surrogate for tree presence-absence data is widespread [@iverson2013tree;@iverson1998predicting], so this complication is a general one faced by forest ecologists and global change ecologists.  In prior comparative analyses of the PLS and FIA data, Goring et al. [-@goring2016preset] tested the effect of differential sampling design on the mapped patterns of novel and disappearing forests, and found little effect.

@@

Historical and modern climate data are derived from monthly PRISM data for 1895 to 2014 CE [@prism2004]. The PRISM climate variables are available at an 800m resolution and have been interpolatedfrom climate stationlocations using elevation, aspectand other data.  Estimates for July temperature ($T_{jul}$), January temperature ($T_{jan}$), the annual temperature range ($T_{diff}$), and annual precipitation ($P_{ann}$) were obtained from PRISM and resampled to the 8 x 8km grid used for PLS and FIA data using bilinear interpolation. Two periods were selected for the calculation of climate normals: 1895 to 1919 for the historical period and 1990 - 2014 for the modern period.  There is an unavoidable temporal mismatch between the historical climate data and the PLS survey period, on the order of 50 to 70 years.  This temporal mismatch is unavoidable; few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatio-temporal resolution for this time period.  The effects of temporal mismatch should be minor given that Northern Hemisphere temperature trends were generally small during the 19^th^ century [@mann2009global].  To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we compare the instrumental record against tree-ring-based paleoclimate reconstructions of the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`).

@@

`r fig_nums("fourPanel", "(a) Diagram of the 2x2 factorial design used to diagnose the effects of historic climate and vegetation change on the observed shifts in tree-climate relationships. We analyzed observed and hypothetical tree-climate relationships for one of four combinations of pre-settlement vegetation (pink fill, left column) or modern vegetation (blue fill, right column) and pre-settlement climate (solid border, top row) or modern climate (dashed border, lower row). (b-e) Representative normalized probability densities for *Larix* for each of the four combinations of climate and vegetation states, plotted relative to $T_{max}$ (b), $T_{diff}$ (c), $T_{min}$ (d), and $P_{ann}$ (e).")`

Distributions of tree genera presences in climate space are described as univariate density functions (PDFs) for both the pre-settlement and modern eras.  Differences among PDFs are quantified using Hellinger distances [@matthes2016benchmarking] to compare PDFs among the four combined states of vegetation and climate ($V_{H}C_{H}$, $V_{H}C_{M}$, $V_{M}C_{H}$, $V_{M}C_{M}$, where V represents vegetation, C represents climate, and M and H represent modern and historical conditions). The total shift ($d_{tot}$) between pre-settlement and modern conditions is calculated as the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ densities (`r fig_nums("fourPanel", display = "cite")`a: top left to bottom right). The effect of historical climate change on shifting climate distributios of tree genera is determined by calculating the Hellinger distance ($d_{C}$) between PLS and FIA tree distributions when both are overlaid on late 19th-century climates ($V_{H}C_{H}$ versus $V_{M}C_{H}$; `r fig_nums("fourPanel", display = "cite")`a; left column). Conversely, the effect of historical vegetation change on tree-climate distributions is determined by calculating the Hellinger distance ($d_{l}$) between the PLSS tree distributions when overlaid on historical vs. modern climates ($V_{H}C_{H}$ versus $V_{H}C_{M}$; Figure 2a; top row).  We develop an attribution index, ($\delta_{d}=d_{c} - d_{l}$), such that negative values of $\delta_{d}$ indicate stronger influence of historical land use on shifting climate distributions, while positive values indicate a stronger influence of historical climate change on shifting distributions.

To assess whether the effects of historical land use and climatic changes on a genera's climatic distribution is compounding, counteracting or confounding we perform a set of t-tests, using the conservative Bonferroni correction to account for multiple tests.  For each taxon and climate variable pair we test whether $V_{H}C_{H}$ and $V_{H}C_{M}$ (attribution to climate change), and $V_{H}C_{H}$ and $V_{M}C_{H}$ (attribution to land-use change) distributions are significantly different.   If both changes are significant and of the same sign, then the shift is defined as compounding.  If both are significant and are in opposite directions then the shift is counteracting.  In the case that either or both changes are insignificant, then the change is defined as confounding.

All analysis makes use of R [@Team2014], RMarkdown [@rmarkdown2016] and R packages `colorspace`, `captioner`, `raster`, `ggplot2`, `mgcv`, `reshape2`, `plyr`, `gridExtra`, `rgdal`, `RColorBrewer`, `analogue`, and `maptools`. All code is hosted on GitHub (http://github.com/PalEON-Project/Composition_Climate). Maps use the public domain Natural Earth Data (http://naturalearthdata.com) products for political and landscape-level features.

@@

Regionally, $T_{Jan}$ increased `r round(clim_change[4], 1)`^o^C since the 1895-1924 normal (`r fig_nums("land_use_clim", display = "cite")`), $T_{Jul}$ declined `r round(clim_change[2], 1)`^o^C.  Because of the large shift in minimum temperatures, $T_{diff}$ decreased by `r round(clim_change[3], 1)`^o^C. Regionally, $P_{ann}$ has increased by `r round(clim_change[1], 1)`mm.  The Palmer Drought Severity Index is closely coherent with $P_{ann}$ trends from 1895 onward and indicates no major climatic trends between 1800 and 1900 (`r fig_nums("land_use_clim", display = "cite")`e), suggesting that early 20^th^ century climates are a reasonable proxy for mid 19^th^ century climates.

@@

`r fig_nums("gain_loss", "Maps showing areas of gain, loss, and continuous presence of tree genera. 'Loss' indicates 8x8km grid cells where PLS data indicated presence of a genus but FIA data do not record presence (light blue). 'Gain' indicates areas where a genus is absent in the PLS data but present in the FIA data (red). 'Continuous presence' represents locations where both FIA and PLS data indicate presence (dark gray). Losses are more common than gains, particularly in the southwestern portion of tree ranges.  Few taxa show gains, although *Populus*, Fraxinus and Tilia show gains of over 10% (Table 2).")`

@@

`r fig_nums("boxplots", "Changes in the realized climate distributions for tree genera in the upper Midwestern United States. In the box-and-whisker plots, the central line indicates the median, box margins indicate 25^th^ and 75^th^ percentiles, whisker tips indicate 5^th^ and 95^th^ percentiles, and dots indicate outliers beyong the 95^th^ percentile. These shifts in realized climate distributions are produced by the joint effects of historical climate change, pathogen-induced tree mortality, and land-use change.")`

The joint effects of historic climate change and land use on tree-climate distributions are illustrated for *Larix* in Figure 2.  For *Larix* we see clear shifts in climate space when vegetation is held constant (from solid to dashed lines), regardless of whether modern or pre-settlement vegetation distributions are used, indicating that historical climate change plays a strong role in shifting the distribution of *Larix* in climate space across all climate variables.  Historical land use also plays a role in shifting distributions, particularly for $T_{max}$, where losses at the highest $T_{max}$ values are visible from pre-settlement to modern (Fig. 2d). However, for *Larix*, land use overall appears to play a smaller role than climate change.

>>>>

```{r setup-knitr, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  comment = " ",
  error = TRUE,
  eval = TRUE,
  cache = TRUE,
  tidy = TRUE

@@

# Effect of historic land-use and climate change on tree-climate relationships in the northern United States

@@

* Simon J Goring: Co-conceived paper, performed analysis, developed figures and co-wrote the paper
* John W Williams: Co-conceived paper, and co-wrote the paper

@@

* Simon J Goring, Department of Geography, University of Wisconsin, Madison, 550 N Park St, Madison WI 53706
* John W Williams, Department of Geography, University of Wisconsin, Madison, 550 N Park St, Madison WI 53706; Center for Climatic Research, University of Wisconsin, Madison, 1225 W Dayton St, Madison WI 53706

@@

* Main Text: 4900
* Text boxes: 0

@@

**Number of Figures, Tables & Text Boxes**: 6 (5 figures, 1 table)

@@

Contemporary forest inventory data are widely used to understand environmental controls on tree species distributions and to project forest responses to climate change, but the stability and representativeness of contemporary tree-climate distributions is poorly understood. Here we show that tree-climate relationships for 15 common tree genera in the northern US have significantly altered over the last two centuries due to historic land use, pathogen outbreaks, and climate change. Realized niches have shifted towards higher minimum temperatures, cooler maximum temperatures, and higher rainfall. A new attribution method implicates both historic climate change and land use in these shifts, with the relative importance varying among genera and climate variables. Most climate/land-use interactions are compounding, in which historic land-use reinforces climatic shifts in species distributions toward cooler or wetter distributions. The prevalence of compounding interactions imply that contemporary-based models of species distributions may underestimate species resilience to climate change.

@@

Contemporary species occurrence databases, and, increasingly, paleohistorical datasets are used to map distributions of species and species traits within environmental space, study underlying processes, and prepare for climate change [@lamanna2014functional;@moritz2013future;@maguire2015modeling;@ivory2016niche;@pearman2008niche]. These distributions of species in geographic and environmental space, combined with future climate scenarios, can be used to assess species exposure and sensitivity to climate change, identify species and habitats of concern, set conservation priorities, and prepare for climate-driven shifts in habitat suitability [@pellatt2012using;@iverson2013tree;@iverson1998predicting;@anderson2013nichemodels;@guisan2013predicting]. Forest inventory datasets, such as the US Forest Inventory Analysis (FIA), are particularly rich observational datasets that have been used to assess early signals of range shifts via the distributions of mature trees and seedlings [@monleon2015evidence;@zhu2012failure] and to parameterize models of current and future tree distributions [@iverson2008estimating;@iverson1998predicting].

Increasingly, the predictive capacity of ecological models is being improved by integrating contemporary observational data with paleohistorical data, when available [@moritz2013future;@maguire2015modeling]. Projections relying only on contemporary observational data may overlook plasticity in species-climate responses [@maiorano2013building], incompletely delimit species fundamental niches [@jackson2000responses], miss responses of species to past no-analog climates [@veloz2012no], or fail to detect effects of anthropogenic land use, and other historical legacies [@svenning2015influence]. Paleoecological records from the last deglaciation show that the time lag of forest responses to abrupt climate change varies from a few years to centuries [@williamsinpress;@webb1986vegetation;@ordonez2013realized;@svenning2015influence], suggesting that tree distributions have yet to fully respond to recent climate changes. In principle, mechanistic dynamic vegetation models [*e.g.*, @leiblein2016modelling] can represent the physiological, demographic, and dispersal processes that cause tree populations to lag climate change.  In practice, however, the parameterization of these models is poorly constrained for processes operating at large temporal and spatial scales [@moorcroft2006close;@matthes2016benchmarking].

Historic land use has substantially affected species distributions and vegetation structure [*e.g.*, @gehrig2007tree] and may cause species distributions to incompletely fill available climates, or otherwise be at disequilibrium [@early2014climatic;@sarmento2013impacts;@ivory2016niche]. Anthropogenic land use is structured and is often biased within climate space, hence, species distributions and protected natural areas often represent a biased selection from the available environmental space [@scott2001nature]. Biasing can confound studies that use contemporary distributional data to understand the environmental controls on species distributions and predict species responses to future climate scenarios [@pyke2004habitat].

Interactions between climate and land-use change can be classified into three types: compounding, confounding and counteracting [modified from @pyke2004habitat]. Compounding effects occur when climatic and land-use changes cause species distributions to shift in the same direction along an environmental gradient.  For example, warming has been compounded by land-use conversion at the lowest elevations in the northern Sierra Nevada Mountains of California, both causing upslope distributional shifts for butterfly species [@forister2010compounded]. Confounding effects may cause little overall change on species distributions, with land-use pressure occurring across the species' environmental gradient. Predictive models for Garry oak (*Quercus garryana*) using only contemporary data [@pellatt2012using] may be confounded by land-use conversion that excludes Garry oak from the deep-soil sites it formerly occupied [@macdougall2004defining;@pellatt2014environmental], but land use has not systematically shifted Garry oak distributions in climate space. Counteracting effects occur when climate and land use change act in opposite directions, narrowing the available geographic and climatic space available for species. For example, bird species in the Sierra Nevadas are expected to move downslope due to increasing precipitation [@tingley2012push], but this downslope migration is limited by land-use conversion in the lowlands.

@@

`r fig_nums("land_use_clim", "*(a-e) Climate change in the Upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions. Modern climate shows lower $T_{max}$, but higher $T_{min}$ relative to early-19^th^ Century normals, along with increasing $P_{ann}$.  PDSI shows strong coherence with $P_{ann}$ in this region. (f) Patterns of land-use change in the upper Midwestern United States, with the 'Tension Zone', differentiating sub-boreal from southern broadleafed forests superimposed. Land use data from the NLCD [@jin2013comprehensive].*")`

Tree distributions and community composition in the upper Midwestern US have been heavily influenced by land use conversion associated with EuroAmerican Settlement in the 19^th^ and 20^th^ Centuries [@goring2016preset;@schulte2007homogenization;@hanberry2012comparison;@rhemtulla2009historical], outbreaks of exotic pathogens [@barnes1979stem], climate changes accompanying the end of the Little Ice Age [@hotchkiss2007response] and current anthropogenic warming [@wicci2011climate]. Modern forests, as represented by FIA surveys [@woudenberg2010forest], show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System (PLS) prior to widespread land-use conversion in the late 1800s [@goring2016preset;@schulte2007homogenization]. Wildland ecosystems have been converted to agricultural and urban land use [@rhemtulla2009historical], particularly in historical prairie and savanna ecosystems (`r fig_nums("land_use_clim", display = "cite")`f).  In northern forested regions, ranges of early successional species have expanded and abundances of late-successional species are reduced due to widespread logging in the 19^th^ and 20^th^ century, and ongoing land management [@schulte2007homogenization;@goring2016preset].  Additionally, the formerly dominant *Ulmus americana* has experienced rangewide population collapses caused by the introduction of exotic pathogens [@barnes1979stem], as did *Castanea dentata*, with other tree species currently threatened (*Tsuga canadensis*, *Fraxinus* spp.)

Here we demonstrate significant shifts in the climatic distributions of tree genera in the upper Midwest over the last 100 years as a result of historic climate change, land use, and pathogen outbreaks, and we diagnose the relative influence of historic climate and land use changes on these shifts.  We use gridded historic and contemporary climate observations (PRISM) and estimates of pre-settlement and modern forest vegetation [@goring2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwestern United States.  We construct four observed and hypothetical sets of tree-climate relationships, using a 2x2 factorial of pre-settlement and modern vegetation and climates.  We develop a new method, based on Hellinger distances among tree-climate distributions, to quantify the total change and attribute the relative importance of historic vegetation and climate change. We assess whether climate and vegetation changes are compounding, counteracting or confounding based on the magnitude and direction of species-climate shifts. We use this framework to discuss the potential impacts of historical land-use and climate change on inferences about species resilience to climate change, as predicted by species-distribution models based soley on modern distributional data.

@@

Estimates of settlement-era forest composition and tree distribution are based on survey data from the PLS [@white1983history;@schulte2001original;@goring2016preset], aggregated to a 64 km^2^grid across the upper Midwestern United States [@goring2016preset]. We briefly review the methods here. PLS data originally were collected on a 1 x 1 mi grid across the Upper Midwest from 1830 to   1910 [@schulte2001original;@almendinger1996minnesota;@white1983history]. At each survey point surveyors noted the closest two to four trees and recorded their common name, distances from survey points, diameters at breast height, and azimuths to each tree. Goring et al. [-@goring2016preset] aggregated this data to an 8x8 km grid, resulting in a gridded data product with an average of 61 survey points per cell, or approximately 120 trees per cell. This process included data cleaning, taxonomic standardization, and the application of spatially varying correction factors [@goring2016preset]. Using this data, we produce presence-absence distributions for each taxon, across the upper Midwest for 15 common tree genera. A few common names, such as "Ironwood" can not be clearly resolved to a single tree genera and so we combine *Ostrya* and *Carpinus*, and *Thuja* and *Juniperus*. 'Presence' is defined as at least one recorded tree in any 8x8 km grid cell and 'range' is considered to be the extent in climate space of all cells with recorded presence for a specific taxon.

Modern forest data are from the US FIA program, which provides detailed inventories of contemporary tree distributions, for all regions with >10% forest cover [@woudenberg2010forest]. The FIA uses a nationally standardized sampling procedure that has evolved over time.  In the Phase 2/3 system, beginning in 1998, each plot consists of four, 7.32m radius, circular sub-plots which are sampled intensively [@woudenberg2010forest].

Comparisons between historical and modern vegetation must address differences in sampling design between the PLS and FIA datasets [@goring2016preset;@kronenfeld2014validating;@kronenfeld2010influence]. Here, the primary solution for maximizing comparability is via aggregation, both spatially, by aggregating records from both datasets to a common scale (8x8 km), and taxonomically by aggregating taxa to the level of genera. Nonetheless, differences persist.  The FIA contains fewer plots per 8 km cell, with more trees per plot than the PLS data and FIA observations are collected only from forested or partially forested landscapes [@woudenberg2010forest], causing many tree taxa to have smaller spatial extents in the FIA than in the PLS datasets. Scattered trees may be present in regions without permanent or temporary FIA plots, complicating the ability to assign 'absence' within the FIA. However, the practice of using FIA data as a surrogate for tree distribution data is widespread [@iverson2013tree;@iverson1998predicting], so these complications are general. In prior comparative analyses of the PLS and FIA data, Goring et al. [-@goring2016preset] tested the effect of differential sampling design on the mapped patterns of novel and disappearing forests, and found little effect.

@@

Historical and modern climate data are derived from monthly PRISM data for 1895 to 2014 CE [@prism2004]. The PRISM climate variables are available at 800m resolution and have been interpolated from station locations using elevation, aspect and other data.  Estimates for July temperature ($T_{max}$), January temperature ($T_{min}$), annual temperature range ($T_{diff}$), and annual precipitation ($P_{ann}$) were resampled to the 8 x 8km grid used for PLS and FIA data. Two periods were selected for calculation of climate normals: 1895 to 1919 for the historical period and 1990 - 2014 for the modern period.  There is an unavoidable temporal mismatch between the historical climate data and the PLS survey period, on the order of 50 to 70 years; few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatio-temporal resolution for this time period.  The effects of temporal mismatch should be minor given that Northern Hemisphere temperature trends were generally small during the 19^th^ century [@mann2009global].  To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we compare the instrumental record against dendroclimatic reconstructions of the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`).

@@

`r fig_nums("fourPanel", "(a) The 2x2 factorial design used to diagnose the effects of historic climate and vegetation change on observed shifts in tree-climate relationships. Observed and hypothetical tree-climate relationships for one of four combinations of pre-settlement vegetation (pink fill, left column) or modern vegetation (blue fill, right column) and pre-settlement climate (solid border, top row) or modern climate (dashed border, lower row). (b-e) Representative normalized probability densities for *Larix* for each of the four combinations of climate and vegetation states, plotted relative to $T_{max}$ (b), $T_{diff}$ (c), $T_{min}$ (d), and $P_{ann}$ (e).")`

Distributions of tree genera within climate space are described as univariate density functions (PDFs) for the pre-settlement and modern eras.  Differences among PDFs are quantified using Hellinger distances [@matthes2016benchmarking] for the four combined states of vegetation and climate ($V_{H}C_{H}$, $V_{H}C_{M}$, $V_{M}C_{H}$, $V_{M}C_{M}$, where V represents vegetation, C represents climate, and M and H represent modern and historical conditions). The total shift ($d_{tot}$) is calculated as the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ (`r fig_nums("fourPanel", display = "cite")`a: top left to bottom right). The effect of historical climate change on shifting climate distributions of tree genera is $d_{C}$, the difference between PLS and FIA tree distributions, overlaid on late 19th-century climates ($V_{H}C_{H}$ versus $V_{H}C_{M}$; `r fig_nums("fourPanel", display = "cite")`a; left column). The effect of historical land use change on tree-climate distributions is $d_{l}$, the difference between the PLSS tree distributions when overlaid on historical vs. modern climates ($V_{H}C_{H}$ versus $V_{M}C_{H}$; Figure 2a; top row).

We develop an attribution index, ($\delta_{d} = d_{c} - d_{l}$), such that negative values of $\delta_{d}$ indicate stronger influence of historical land use on shifting climate distributions, while positive values indicate a stronger influence of historical climate change on shifting tree distributions.

To assess whether the effects of historical land use and climatic changes are compounding, counteracting or confounding we perform a set of t-tests, using the conservative Bonferroni correction to account for multiple tests.  For each taxon and climate variable pair we test whether $V_{H}C_{H}$ and $V_{H}C_{M}$ (attribution to climate change), and $V_{H}C_{H}$ and $V_{M}C_{H}$ (attribution to land-use change) distributions are significantly different.   If both changes are significant and of the same sign, then the shift is compounding.  If both are significant and in opposite directions then the shift is counteracting.  If either or both changes are not significant, then the change is defined as confounding.

All analysis uses R [@Team2014], RMarkdown [@rmarkdown2016] and R packages `colorspace`, `captioner`, `raster`, `ggplot2`, `mgcv`, `reshape2`, `plyr`, `gridExtra`, `rgdal`, `RColorBrewer`, `analogue`, and `maptools`. All code is hosted on GitHub (http://github.com/PalEON-Project/Composition_Climate). Maps use the public domain Natural Earth Data (http://naturalearthdata.com) products for political and landscape features.

@@

Regionally, $T_{min}$ increased `r round(clim_change[4], 1)`^o^C since the 1895-1924 normal (`r fig_nums("land_use_clim", display = "cite")`), $T_{max}$ changed by `r round(clim_change[2], 1)`^o^C, while $T_{diff}$ changed by `r round(clim_change[3], 1)`^o^C. Regionally, $P_{ann}$ has increased by `r round(clim_change[1], 1)`mm.  The Palmer Drought Severity Index is closely coherent with $P_{ann}$ trends from 1895 onward and indicates no major climatic trends between 1800 and 1900 (`r fig_nums("land_use_clim", display = "cite")`e), suggesting that early 20^th^ century climates are a reasonable proxy for mid-19^th^ century climates.

@@

`r fig_nums("gain_loss", "Maps showing areas of gain, loss, and continuous presence of tree genera. 'Loss' indicates 8x8km grid cells where PLS data indicated presence of a genus but FIA data do not record presence (light blue). 'Gain' indicates areas where a genus is absent in the PLS data but present in the FIA data (red). 'Continuous presence' represents locations where both FIA and PLS data indicate presence (dark gray). Losses are more common than gains, particularly in the southwestern portion of tree ranges.  Few taxa show gains, although *Populus*, Fraxinus and Tilia show gains of over 10% (Supplemental Material).")`

@@

`r fig_nums("boxplots", "Changes in the realized climate distributions for tree genera in the upper Midwestern United States. In the box-and-whisker plots, the central line indicates the median, box margins indicate 25^th^ and 75^th^ percentiles, whisker tips indicate 5^th^ and 95^th^ percentiles, and dots indicate outliers beyond the 95^th^ percentile. These shifts in realized climate distributions are produced by the joint effects of historical climate change, pathogen-induced tree mortality, and land-use change.")`

The joint effects of historic climate change and land use on tree-climate distributions are illustrated for *Larix* (Figure 2).  For *Larix* we see clear shifts in climate space when vegetation is held constant (from solid to dashed lines), regardless of whether modern or pre-settlement vegetation distributions are used, indicating that historical climate change plays a strong role in shifting the distribution of *Larix* in climate space across all climate variables.  Historical land use contributes to shifting distributions, particularly for $T_{max}$, where losses at the highest $T_{max}$ values are visible from pre-settlement to modern (Fig. 2d). However, for *Larix*, land use overall appears to exert a weaker influence than climate change.
